Low Complexity Adaptive Sonar Imaging

### 2018-02-06 LCAforActiveSonarImaging/lca_submit1.tex
Major commit before messing with PhD thesis

\newcommand\mediaPath{gfx/media}
\newcommand\mediaI{\multimedia{\mediaPath/media1.mp4}{Media Movie 1}}
\newcommand\mediaII{\multimedia{\mediaPath/media2.mp4}{Media Movie 2}}

@@

We show that a robust MVDR implementation typically creates responses ranging between that of the rectangular and Hamming. We let LCA select from a set of Kaiser windows with responses in this range, and add some steered variations of each. We limit the steering to roughly half the -3\,dB width of the window's amplitude response. Using experimental data from the Kongsberg Maritime HISAS 1030 sonar we find that LCA and MVDR produce nearly identical images of large scenes, both being superior to DAS. On point targets LCA is able to double the resolution compared to DAS, or provide half that of MVDR. This performance is achieved with a total of 6 windows; the rectangular and Kaiser window with $\beta=5$, in unsteered version, and left and right steered to the steering limit. Slightly smoother images are produced if the window count is increased to 15, but past this we observe minimal difference. Finally we show that LCA work just as well if Kaiser windows are substituted with trigonometric ones.

All our observations and experiences point to LCA being very easy to manage. It simply works, and is surprisingly insensitive to the exact type of window function, steering amount, or number of windows. It can be efficiently implemented on parallel hardware, and handles any scene without the need for parameter adjustments.

@@

\IEEEPARstart{T}{he} best adaptive sonars are found in nature. Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats. Active human made sonars have little to no adaptability on transmit, but several adaptive methods have been investigated post process the received data. Perhaps the best studied method for reconstructing a sonar image is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}. It adapts the sonar array's response to minimize the influence of noise and interference in the final image.

In many cases MVDR can improve the contrast and resolution of active sonar images compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}. However, while the computational complexity of conventional beamformers are linear with the number of channels, O($M$), MVDR is at O($M^3$). This makes MVDR impracticable and sometimes infeasible. We have shown in previous studies that MVDR's complexity can be reduced, and its implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}. However, while MVDR relies on estimating and inverting a spatial covariance matrix, there exist a method that avoid this altogether.\todo{samme avsnitt?}

This method is called the Low Complexity Adaptive (LCA) beamformer. It was first proposed by Synnev\aa{}g in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar results to MVDR in a system with focused transmit beams. LCA applies a set of predefined windows and selects the one offering the best noise suppression. It relies on the same optimization criterion as MVDR, and may be considered a version of MVDR with a reduced window solution space. Alternatively, it may be viewed as a delay-and-sum (DAS) beamformer with an adaptive extension.

Synnev\aa{}g used a window set comprised of rectangular, Kaiser and inverted Kaiser functions\footnote{May also be called Kaiser-Harris windows.}. He made LCA choose from 12 unique windows, 6 of which were micro-steered variants of a fairly wide Kaiser window. While achieving good results, it was not made clear how LCA performs with more or fewer windows, and how steering affects it all.

We use LCA with variations of the Kaiser window, but a comparison to using trigonometric windows will also be provided. We extend Synnev\aa{}g's work by steering all our windows and not just some, and elaborate on the type of Kaiser windows to include and the method of steering. Then the method is applied to experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), with an operating frequency of 100\,kHz and transmit angle of 25$^\circ$.

Our results show that MVDR computes weights with responses that tend to be symmetric and steered within a fraction of the window's -\,3dB resolution. When letting LCA choose from Kaiser windows with similar responses, we obtain images with improved noise suppression and resolution. The resolution gain is made predictable by limiting the maximum steering to a fraction of the window's -3\,dB width. Overall, LCA produces images that are similar to MVDR, but with a resolution in between that of DAS and MVDR. LCA is also inherently robust, easy to implement and fairly easy to understand.

This article is outlined as follows: In Section \ref{sec:beamforming} we offer a gentle introduction to beamforming. Then we move on to adaptive beamforming and describe the MVDR and LCA methods \ref{sec:mvdr} and \ref{sec:lca}, respectively. We use LCA with the Kaiser window which we describe in \ref{sec:lca_kaiser_windows}, explain how we steer it in \ref{sec:lca_steering}, and by how much in \ref{sec:lca_steering_bounds}. Finally we add some remarks on using the trigonometric window instead of Kaiser in \ref{sec:lca_trigonometric}, and of of sampling considerations in \ref{sec:lca_oversampling}. Results and discussion a provided in Section \ref{sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{sec:results_lca_window_function}, assess the range of window types and steering in \ref{sec:results_window_parameters}, determine the number of windows needed in \ref{sec:results_database_size}, and add a note on computational complexity in \ref{sec:results_complexity}. Finally we conclude in \ref{sec:conclusion}.

@@

To form a sonar image we need to estimate source locations and amplitudes. For this purpose we apply a spatial bandpass filter to the backscattered wavefield data. The filter is called an array processor, or receive beamformer. The basic principle is to apply delay and weights to the sensor channels before summing them up, chosen such that signals from the location of interest are summed coherently, while other sources sum incoherently.

Assume that the wavefield has been sampled by an $M$ element uniform linear array, and that the signal signature has been removed by a matched filter. Let $x_m[\theta,n]$ be the delayed data from the $m$th channel, where the $\theta$ and $n$ is the azimuth angle and range sample of the focus point, respectively. Each angle $\theta$ will be processed independently, so to simplify notation we will assume the dependence on $\theta$ to be implicit from now on.

The output $z[n]$ of a beamformer is defined as the weighted sum of all the delayed data samples:
%
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{eq:beamformer_output}
\end{align}
%
where $w_m$ is the weight factor assigned to channel $m$. A weight set is commonly called a window or a taper function. Applying a window that trails off towards the edges improves noise suppression at the cost of reduced resolution~\cite{Harris1978}. Real windows lead to symmetric responses, while complex weights allow asymmetric responses.

@@

MVDR seeks to minimize the output power of the beamformer, under the constraint of a unity gain in some desired direction $\phi$~\cite{Capon1969}:
%
\begin{align}
\underset{\w[n]}{\argmin}\, E\{\big|z[n]\big|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\nn
\text{subject to } \w[n]\a_\phi &= 1.\label{eq:mvdr_definition}
\end{align}
%
Here $\a_\phi$ is a steering vector, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x\x\H\}$ is the spatial covariance matrix. This is a convex optimization problem with the solution:
%
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi},\label{eq:mvdr_weights}
\end{gather}
where $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. The problem lies in estimating and inverting this spatial covariance matrix. We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}, for true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}, and to improve robustness to parameter errors we add $d$ percent diagonal loading~\cite{Cox1987,Maksym1979}. These steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.

\begin{figure}[tbp]%
\includegraphics[width=\linewidth]{gfx/calc_kaiser_3dB.eps}%
\caption{Each Kaiser window is steered in the interval $\phi=[0, \phi_{\text{-3dB}}(\beta)]$. The angle $\phi_{\text{-3dB}}(\beta)$ is the amount of steering needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window. With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.}\label{windows_steering}
\end{figure}

\begin{figure}[tbp]%
\includegraphics[width=\linewidth]{gfx/cross_tagged.png}%
\caption{One of the objects used to test the beamformers is a 1\,m by 1\,m cross attached to an anchor with a diameter of approximately 13\,cm. Source image curtesy of Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).}\label{cross}
\end{figure}

@@

\begin{figure*}[t]%
\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds.pdf}%
\caption{\emph{Determining Kaiser parameter boundaries}: LCA images created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied. The upper left image is equal to DAS with a rectangular window. For all images we measure the lateral distance $\Delta x$ between the two -3dB points of the anchor. These are specified relatively to the reference distance  $\Delta x_\text{ref}$ from the upper left DAS image.
\newline
\emph{Upper bound $\beta$:}\hfill
\parbox[t]{.89\linewidth}{Observe that LCA gets better at suppressing sidelobes in the image as we increase the upper bounds for $\beta$. \Fig{mvdr_selected_windows} suggests that MVDR tend to choose windows with levels lower than that of Kaiser with $\beta=2$, but here we observe further improvement going to $\beta=5$.}\newline
\emph{Upper bound $\phi$:}\hfill
\parbox[t]{.89\linewidth}{As we increase the upper bound of the steering $\phi$, we also increase the lateral image resolution. }
}\label{oversampling_mosaic_bounds}
\end{figure*}

@@

In practice, we estimate the beamformer output power as
%
\begin{align*}
E\big\{\big|z[n]\big|^2\big\} = \frac{1}{N_k} \sumb{n'=n-K}{n+K} \big| z[n'] \big|^2
\end{align*}
%
which computes the sample power average over $N_k=2K+1$ temporal samples. This is the same temporal averaging method we use for MVDR. In our case the matched filtered signal is a just few samples long, hence $K=1$ will be used for both MVDR and LCA throughout this work.

\todo{Drop this paragraph altogether?}If extra flexibility is desired, we can also estimate the variance by a weighted combination of local pixels
%
\begin{align}
E\big\{\big|z[x,n]\big|^2\big\} = \frac{1}{N_x N_k} \sumb{\nmat{\scriptstyle x'=\\[-2mm]\scriptstyle x-X}}{x+X} \sumb{\nmat{\scriptstyle n'=\\[-2mm]\scriptstyle n-K}}{n+K} w[x',n']\big| z[x',n'] \big|^2
\end{align}
%
where $N_x = 2X+1$ is the number of azimuth lines to average over, and $w_\text{prf}[x',n']$ is a normalized 2 dimensional weight function. 
% 
% \begin{itemize}
% \item \emph{Window function}
% \item \emph{Parameter boundaries}
% \item \emph{Number of windows needed}
% \item \emph{Oversampling need}
% \item \emph{Image quality}
% \end{itemize}

@@

As will be demonstrated in upcoming sections, the LCA beamformer works very well with windows generated from the Kaiser-Bessel function. We will express it in vector form as:
%
\begin{align}
\f_\beta = \bmat{
f_0(\beta) \\
\vdots\\
f_{M-1}(\beta)
}\label{eq:kaiser_window_vector}
\end{align}
%
where
%
\begin{align}
f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}\label{eq:kaiser_window_element}
\end{align}
%
and $I_0$ is the zeroth order modified Bessel function of the first kind:
%
\begin{align}
I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.\label{eq:modified_bessel_first_kind}
\end{align}
%
The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given time-bandwidth product specified with the parameter $\beta$:
%
\begin{align}
\beta = \frac{TB}{2},
\end{align}
%
where $T$ is the extent of the window in time\todo{Should perhaps say space-bandwidth product?} and $B$ is its bandwidth. Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level. When $\beta=0$ the window becomes rectangular, while at large values ($\beta>5$) the window converges to a Gaussian both in time and frequency. This class of windows is generally considered well suited for separating closely spaced\todo{better word?} sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.

@@

To test the performance of the MVDR and LCA beamformers, we have processed data acquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2009}. It is a high resolution synthetic aperture sonar with 1.2\;m array length, 100\;kHz operating frequency, 40\;kHz bandwidth and 25$^\circ$ opening angle on transmission.

Two different scenes will be studied. One of the 1500 DWT oil tanker wreck Holmengraa lying at a slanted seabed at 77\;m depth outside of Horten, Norway~\cite{holmengraa}. It measures 68\;m by 9\;m and fills most of the highlighted sector. This data was collected by the Norwegian Defense Research Establishment and Kongsberg Maritime. The other scene is of an 1\,m by 1\,m iron cross, connected to a an anchor with 13\,cm diameter (\Fig{cross}). This data was collected by the Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71). Both scenes are imaged in sectorscan mode.

In the image reconstruction we have run MVDR with subarray length $L=16$ and $d=1\%$ diagonal loading\todo{also used for element distance}. Both MVDR and LCA was run with $K=1$ temporal averaging. This is a fairly aggressive yet stable set of parameters.\todo{suitable ref?}

@@

LCA iterates through a set of windows in search of the one that minimizes the noise and interference power in each image pixel. It can be viewed as either a reduced solution space version of MVDR, or as an multi-taper extension to DAS. It creates images similar to MVDR, but at a fraction of the computational cost. It is also inherently robust and insensitive to the particular set of windows it selects from, supported by the observation that LCA can be used with either Kaiser or trigonometric windows and produce near identical images.

When using LCA with Kaiser windows we made their responses similar to those typically chosen by a robust MVDR in key areas in the image. We suggest an upper bound for $\beta$ between 2 and 5, where 5 is slightly better at suppressing sidelobes. For $\phi$ we suggest to steer either to 50\% or 100\% of the \minus{}3\,dB width, which leads to a resolution of 73\% or 54\% of the rectangular window, respectively. As lower boundaries we set $\beta=0$ to always include a rectangular window, and $\phi=0$ to always include the centered window.

The tendency of the LCA to perform similarly to the MV beamformer seems to indicate that a full solution space is rarely needed in real world scenarios.

% 
% LCA is largely insensitive to the choice of windows and steering. 
% LCA works well for all the boundaries
% Variables of interest:

@@

@@

@@

\bibliographystyle{IEEEtran}
\bibliography{references}
% 
% 
\input{bio/jo_inge_buskenes}
\input{bio/roy_edgar_hansen}
\input{bio/andreas_austeng}

>>>>

% \newcommand\mediaPath{gfx/media}
\newcommand\mediaPath{http://folk.uio.no/joibu/media/2015_JOE_LCA}

\newcommand\mediaI{\multimedia{\mediaPath/media1.mp4}{Media Movie~1}}
\newcommand\mediaII{\multimedia{\mediaPath/media2.mp4}{Media Movie~2}}

@@

We show that a robust MVDR implementation typically creates windows spanning between the rectangular and Hamming. We let LCA select from a set of Kaiser windows with responses in this span, and add some steered variations of each. We limit the steering to roughly half the -3\,dB width of the window's amplitude response. Using experimental data from the Kongsberg Maritime HISAS 1030 sonar we find that LCA and MVDR produce nearly identical images of large scenes, both being superior to DAS. On point targets LCA is able to double the resolution compared to DAS, or provide half that of MVDR. This performance is achieved with a total of 6 windows; the rectangular and Kaiser window with $\beta=5$, in unsteered version, and left and right steered to the steering limit. Slightly smoother images are produced if the window count is increased to 15, but past this we observe minimal difference. Finally we show that LCA works just as well if Kaiser windows are substituted with trigonometric ones.

All our observations and experiences point to LCA being very easy to understand and manage. It simply works, and is surprisingly insensitive to the exact type of window function, steering amount, or number of windows. It can be efficiently implemented on parallel hardware, and handles any scene without the need for parameter adjustments.

@@

\IEEEPARstart{T}{he} best adaptive sonars are found in nature. Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats. Active human made sonars have little to no adaptability on transmit, but several methods have been investigated to adaptively post process the received data. Perhaps the best studied method for reconstructing a sonar image is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}. It adapts the sonar array's spatial response to minimize the influence of noise and interference in the final image.

In many cases MVDR can improve the contrast and resolution of active sonar images compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Lo2004}. However, while the computational complexity of conventional beamformers are linear with the number of channels, O($M$), MVDR is at O($M^3$). This is because MVDR relies on estimating and inverting a covariance matrix. In sonar systems with less than 32 channels the complexity can be reduced and the implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}. For larger systems the inversion step dominates the computation. This can be dealt with by approximating the full covariance matrix with a small one using space reduction techniques such as beamspace processing~\cite{Asen2013}, or the closely related principal component analysis (PCA) method~\cite{Kim2014}. Another alternative is to assume spatial stationarity to form more easily invertible Toepliz matrices~\cite{Asl2012}. However, the  mentioned methods are still relatively slow compared to DAS.

A much faster and simpler alternative to MVDR is the Low Complexity Adaptive (LCA) beamformer. Based on an idea by Vignon~\cite{Vignon2008}, it was first introduced by Synnev\aa{}g \emph{et al}. in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams. LCA applies a set of predefined windows, or apodization weights, and selects the one offering the best sidelobe suppression. It relies on the same optimization criterion as MVDR, thus may be considered a version of MVDR with a reduced window solution space. Alternatively, it may be viewed as a multi-apodization technique. One such method is described by Stankwitz \emph{et al}. in radar imaging~\cite{Stankwitz1995}, where the best out of 2 or 3 windows are selected. However, LCA is different in that it typically selects from a larger pool of windows and allows micro-steered variations of them. This gives LCA an adaptive freedom that is somewhere in between that of traditional multi-apodization techniques and MVDR.

Synnev\aa{}g \emph{et al}. suggested using a window set comprised of rectangular, Kaiser (or Kaiser-Harris) and inverted Kaiser functions. In total he made LCA choose from 12 unique windows, 6 of which were micro-steered variants of a fairly wide Kaiser window, and with the rectangular window included. This was motivated by the desire to use wide windows with maximum sidelobe suppression near the receiver, and narrow windows at greater depths for maximum sensitivity and depth penetration. While achieving good results, it was not made clear how how many windows LCA needs, or how they ideally should be steered. Also, it is not apparent how LCA performs in sonar systems where the transmit beam is unfocused.

In this study we expand our previous work of applying LCA to active sonar imaging~\cite{Buskenes2011}. We use variations of the parametric Kaiser window, but also compare it to the trigonometric window function. We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by the type and number of windows included. Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), which operates at 100\,kHz with an element beamwidth of 23$^\circ$.

Our results show that when MVDR is modified to work in active sonar imaging, it computes weights with spatial responses that tend to be symmetric and steered within a fraction of the window's -\,3dB resolution. When letting LCA choose from Kaiser windows with similar responses, we obtain images with improved noise suppression and resolution. The resolution gain is made predictable by limiting the maximum steering to a fraction of the window's -3\,dB width. Overall, LCA produces images that are similar to MVDR, but with a resolution in between that of DAS and MVDR. LCA is also inherently robust, easy to implement, and fairly easy to understand.

This article is outlined as follows: In Section \ref{sec:beamforming} we offer a gentle introduction to beamforming. Then we move on to adaptive beamforming and describe the MVDR and LCA methods in \ref{sec:mvdr} and \ref{sec:lca}, respectively. We use LCA with the Kaiser window which we describe in \ref{sec:lca_kaiser_windows}, explain how we steer it in \ref{sec:lca_steering}, and by how much in \ref{sec:lca_steering_bounds}. Finally we add some remarks on using the trigonometric window instead of Kaiser in \ref{sec:lca_trigonometric}, and of of sampling considerations in \ref{sec:lca_oversampling}. Results and discussion are provided in Section \ref{sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{sec:results_lca_window_function}, assess the range of window types and steering in \ref{sec:results_window_parameters}, determine the number of windows needed in \ref{sec:results_database_size}, and add a note on computational complexity in \ref{sec:results_complexity}. Section \ref{sec:conclusion} provides a conclusion.

@@

A sonar image is formed by estimating source locations and amplitudes. For this purpose a spatial bandpass filter is applied to the backscattered wavefield data. The filter is called an array processor or receive beamformer. The basic principle is to apply delay and weights to the sensor channels before summing them up, chosen such that signals from the location of interest are summed coherently, while other sources sum incoherently.

Assume that the wavefield has been sampled by an $M$ element uniform linear array, and that the signal signature has been removed by a matched filter. Let $x_m[\theta,n]$ be the delayed data from the $m$th channel, where the $\theta$ and $n$ are the azimuth angle and range sample of the focus point, respectively. Each angle $\theta$ will be processed independently, so to simplify notation we will assume the dependence on $\theta$ to be implicit from now on.

The output $z[n]$ of a beamformer is defined as the weighted sum of all the delayed data samples:
%
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{eq:beamformer_output}
\end{align}
%
where $w_m$ is the weight factor assigned to channel $m$. A weight set is commonly called a window, apodization or taper function. Applying a window that trails off towards the edges improves noise suppression at the cost of reduced resolution~\cite{Harris1978}. Real windows lead to symmetric responses, while complex weights allow asymmetric responses.

@@

MVDR seeks to minimize the output variance of the beamformer, under the constraint of a unity gain in some desired direction $\phi$~\cite{Capon1969}. With the assumption of zero mean data this can be formulated as: 
%
\begin{align}
\underset{\w[n]}{\argmin}\, E\{\big|z[n]\big|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\nn
\text{subject to } \w^H[n]\a_\phi &= 1.\label{eq:mvdr_definition}
\end{align}
%
Here $\a_\phi$ is a steering vector, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x[n]\x\H[n]\}$ is the spatial covariance matrix for the full array. This is a convex optimization problem with the solution:
%
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi}.\label{eq:mvdr_weights}
\end{gather}
The problem lies in estimating and inverting this spatial covariance matrix. We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}, for true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}, and to improve robustness to parameter errors we add {\large$\epsilon$} percent diagonal loading~\cite{Cox1987,Maksym1979}. These steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.

\begin{figure}[tbp]\centering%
\ifPeerReview%
\includegraphics[width=.5\linewidth]{gfx/calc_kaiser_3dB.eps}%
\else%
\includegraphics[width=\linewidth]{gfx/calc_kaiser_3dB.eps}%
\fi%
\caption{Each Kaiser window is steered in the interval $\phi=[0, \phi_{\text{-3dB}}(\beta)]$. The angle $\phi_{\text{-3dB}}(\beta)$ is the amount of steering needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window. With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.}\label{windows_steering}
\end{figure}

\begin{figure}[tbp]\centering%
\ifPeerReview%
\includegraphics[width=.5\linewidth]{gfx/cross_tagged.png}%
\else%
\includegraphics[width=\linewidth]{gfx/cross_tagged.png}%
\fi%
\caption{One of the objects used to test the beamformers is a 1\,m by 1\,m resolution test cross attached to an anchor with a diameter of approximately 13\,cm. Source image curtesy of Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).}\label{cross}
\end{figure}

@@

\begin{figure*}[t]%
\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds.pdf}%
\caption{\emph{Determining Kaiser parameter boundaries}: LCA images of the resolution test cross and anchor created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied. The upper left image is equal to DAS with a rectangular window. For all images we measure the lateral distance $\Delta x$ between the two -3dB points of the anchor (red object). These are specified relatively to the reference distance  $\Delta x_\text{ref}$ from the upper left DAS image.
\newline
\emph{Upper bound $\beta$:}\hfill
\parbox[t]{.89\linewidth}{Observe that LCA gets better at suppressing sidelobes in the image as we increase the upper bounds for $\beta$. \Fig{mvdr_selected_windows} suggests that MVDR tend to choose windows with levels lower than that of Kaiser with $\beta=2$, but here we observe further improvement going to $\beta=5$.}\newline
\emph{Upper bound $\phi$:}\hfill
\parbox[t]{.89\linewidth}{As we increase the upper bound of the steering $\phi$, we also increase the lateral image resolution. }
}\label{oversampling_mosaic_bounds}
\end{figure*}

@@

In practice, we estimate the beamformer output power by computing a sample power average $s_z^2$:
%
\begin{align*}
E\big\{\big|z[n]\big|^2\big\} \approx s_z^2 = \frac{1}{N_k} \sumb{n'=n-K}{n+K} \big| z[n'] \big|^2
\end{align*}
%
which computes the sample power average over $N_k=2K+1$ temporal samples. This is the same temporal averaging method we use for MVDR. In our case the bandwidth of our matched filtered signal is a just few samples long, hence $K=1$ will be used for both MVDR and LCA throughout this work.

Since pixels are correlated laterally, we can include a weighted combination of these as well to improve the variance estimation:
%
\begin{align}
s_z^2 = \frac{1}{N_x N_k} \sumb{\nmat{\scriptstyle x'=\\[-2mm]\scriptstyle x-X}}{x+X} \sumb{\nmat{\scriptstyle n'=\\[-2mm]\scriptstyle n-K}}{n+K} \omega[x',n']\big| z[x',n'] \big|^2
\end{align}
%
where $N_x = 2X+1$ is the number of azimuth lines to average over, and $\omega[x',n']$ is a normalized 2 dimensional weight function. However, we observed no visual improvement doing this, and the effect of micro-steered windows is hard to predict. Henceforth no lateral averaging will be applied.
% 
% \begin{itemize}
% \item \emph{Window function}
% \item \emph{Parameter boundaries}
% \item \emph{Number of windows needed}
% \item \emph{Oversampling need}
% \item \emph{Image quality}
% \end{itemize}

@@

As will be demonstrated in upcoming sections, the LCA beamformer works very well with windows generated from the Kaiser-Bessel function. We will express it in vector form as:
%
\begin{align}
\f_\beta = \bmat{
f_0(\beta) \\
\vdots\\
f_{M-1}(\beta)
}\label{eq:kaiser_window_vector}
\end{align}
%
where
%
\begin{align}
f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}\label{eq:kaiser_window_element}
\end{align}
%
and $I_0$ is the zeroth order modified Bessel function of the first kind:
%
\begin{align}
I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.\label{eq:modified_bessel_first_kind}
\end{align}
%
The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given space-bandwidth product related to the Kaiser parameter $\beta$ as:
%
\begin{align}
\beta = \frac{TB}{2},
\end{align}
%
where $T$ in our case is the spatial extent of the window and $B$ is its bandwidth. Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level. When $\beta=0$ the window becomes rectangular, while at large values ($\beta>5$) the window converges to a Gaussian both in time and frequency. This class of windows is generally considered well suited for separating closely spaced sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.

@@

To test the performance of the MVDR and LCA beamformers, we have processed data acquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2011}. It is a high resolution synthetic aperture sonar with 1.2\;m array length, 100\;kHz operating frequency, 30\;kHz bandwidth and 23$^\circ$ receiver element beamwidth.

Two different scenes will be studied. One of the 1500 DWT oil tanker wreck Holmengraa lying at a slanted seabed at 77\;m depth outside of Horten, Norway~\cite{holmengraa}. It measures 68\;m by 9\;m and fills most of the highlighted sector. This data was collected by the Norwegian Defense Research Establishment and Kongsberg Maritime. The other scene is of an 1\,m by 1\,m iron resolution test cross, connected to a an anchor with 13\,cm diameter (\Fig{cross}). This data was collected by the Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71). Both scenes are imaged in sectorscan mode.

In the image reconstruction we have run MVDR with subarray length $L=16$ and $\text{\large$\epsilon$}=1\%$ diagonal loading. Both MVDR and LCA was run with $K=1$ temporal averaging. This is a fairly aggressive yet stable set of parameters~\cite{Synnevag2009a}.

@@

LCA seeks to improve image resolution and contrast by minimizing the power of noise and interference in each image pixel. This is the same optimization criterion as MVDR uses, but instead of computing an optimal array window like MVDR does, it selects the best window out of a predefined set. Hence, it can be viewed as either a reduced and discrete window space version of MVDR, or as an advanced multi-apodization technique. What we have studied in this paper is which windows we should ideally let LCA choose from.

We have found the LCA to be surprisingly insensitive to the exact window function it is used with, supported by the observation that near identical images were produced whether we used Kaiser or trigonometric windows. However, it needs to be able to choose between wide and narrow window responses, since the wide ones offer the best sidelobe suppression and the narrow ones offer the best sensitivity. We achieved good results using 2-3 Kaiser windows in the range $\beta\in[0,5]$, and observed minimal difference when adding more windows or adjusting the upper bound of $\beta$. For the steering angle we suggest a value in the range $\phi\in[0,\pm\delta\phi_\text{3dB}(\beta)]$, where $\phi_\text{3dB}(\beta)$ is the bandwidth of the respective window, and $\delta$ is a scaling parameters. Setting $\delta$ to either 50\% or 100\% reduces the lateral point target size to 73\% or 53\% compared to the rectangular window, respectively. Hence, this parameter can be used to control the aggressiveness and resolution gain of LCA.

In summary, LCA is an attractive alternative to other adaptive beamformers due to being very fast, simple to understand, practically parameter-free, inherently robust, and able produce images similar to that of MVDR. Contrary to MVDR, it has a low computational complexity and can be easily and efficiently accelerated using e.g. GPUs. Since LCA performs just as well if we used trigonometric windows instead of Kaiser, we can even compute the optimal window parameter analytically for any given steering angle, and avoid the search over the window parameter entirely.

% The tendency of LCA to produce images of similar quality as MVDR seems to indicate that a full solution space is rarely needed in real world scenarios.

%  It initially very is also very fast, and is very well suited for acceleration using  very fast, easy to implement even on parallel hardware,

% Its computational complexity is only of O($N_wM$), and its data parallel nature makes it very well suited for acceleration using e.g. GPUs. compared to MVDR's O($M^3$).
% 
% LCA is largely insensitive to the choice of windows and steering. 
% LCA works well for all the boundaries
% Variables of interest:

@@

\section{Steering angle}

@@

\section{Multimedia Files and Source Code}\label{source_code}

@@

\ifBuildBibliography
   \bibliographystyle{IEEEtran}
   \bibliography{references}

\else

% Generated by IEEEtran.bst, version: 1.13 (2008/09/30)
\begin{thebibliography}{10}
\providecommand{\url}[1]{#1}
\csname url@samestyle\endcsname
\providecommand{\newblock}{\relax}
\providecommand{\bibinfo}[2]{#2}
\providecommand{\BIBentrySTDinterwordspacing}{\spaceskip=0pt\relax}
\providecommand{\BIBentryALTinterwordstretchfactor}{4}
\providecommand{\BIBentryALTinterwordspacing}{\spaceskip=\fontdimen2\font plus
\BIBentryALTinterwordstretchfactor\fontdimen3\font minus
  \fontdimen4\font\relax}
\providecommand{\BIBforeignlanguage}[2]{{%
\expandafter\ifx\csname l@#1\endcsname\relax
\typeout{** WARNING: IEEEtran.bst: No hyphenation pattern has been}%
\typeout{** loaded for the language `#1'. Using the pattern for}%
\typeout{** the default language instead.}%
\else
\language=\csname l@#1\endcsname
\fi
#2}}
\providecommand{\BIBdecl}{\relax}
\BIBdecl

\bibitem{Capon1969}
J.~Capon, ``{High-resolution frequency-wavenumber spectrum analysis},''
  \emph{Proceedings of the IEEE}, vol.~57, no.~8, pp. 1408--1418, 1969.

\bibitem{Blomberg2013}
A.~E.~A. Blomberg, A.~Austeng, R.~E. Hansen, and S.~A.~V. Synnes, ``{Improving
  Sonar Performance in Shallow Water Using Adaptive Beamforming},'' \emph{IEEE
  Journal of Oceanic Engineering}, vol.~38, no.~2, pp. 297--307, Apr. 2013.

\bibitem{Blomberg2012a}
A.~E.~A. Blomberg, A.~Austeng, and R.~E. Hansen, ``{Adaptive Beamforming
  Applied to a Cylindrical Sonar Array Using an Interpolated Array
  Transformation},'' \emph{IEEE Journal of Oceanic Engineering}, vol.~37,
  no.~1, pp. 25--34, Jan. 2012.
